---
title: 'Data visualisation - part 3: compound figures with cowplot'
author: "Hannah Meyer"
date: "01/08/2020"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 3
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
bibliography: bibliography.bib
---

# Background
There are many add on packages for `ggplot2`. Here, I want to highlight and show
examples for one of those packages: `cowplot` (short for **C**laus **O**.
**W**ilkes plot package, initially developed for Claus's lab members).
It is extensively documented on this webpage:
[cowplot](https://wilkelab.org/cowplot/index.html).

`cowplot` is described as

    provid[ing] various features that help with creating publication-
    quality figures, such as a set of themes, functions to align plots and
    arrange them into complex compound figures, and functions that make it easy
    to annotate plots and or mix plots with images.



# Setting up
In the following two chunks, we will again set up our analys document by 
specifying the options for `knitr` and
```{r setup}
knitr::opts_chunk$set(echo = TRUE,
                      comment = "#>",
                      collapse = TRUE,
                      fig.width = 6,
                      fig.align = "center",
                      out.width = "70%")
```

load all libraries required for our analysis:
```{r libraries, message=FALSE}
library("tidyverse")
library("cowplot")
library("sf")
library("rnaturalearth")
library("RColorBrewer")
```

# Individual visualisations
In the following section, we will load the dataset we worked with in the
previous session and recreate the visualisations we used to show the antigenic
maps, histogram and boxplots of the clusters by time.

Unlike the previous session, where we immediately displayed the output
of each `ggplot2` call, we will now save each plot into a different object.
To do so, we will first create the common `ggplot` object `p`, that relies
on our dataset `coord`:

```{r}
coord <- read_csv("data/2004_Science_Smith_data.csv")
p <- ggplot(data=coord)
```

To recreate the previous plots, we save them into a new object by adding the
appropriate layers to `p`:
```{r}
antigenic_map <- p + 
  geom_point(aes(x=x.coordinate, y=y.coordinate, color=cluster)) +
  scale_color_brewer(type="qual", palette = "Set3") +
  labs(x="Dimension 1 [AU]",
       y="Dimension 2 [AU]",
       color="Cluster") +
  theme_bw()

time_barplot <- p + geom_bar(aes(x=year, fill=cluster),
                             position=position_dodge(preserve="single")) +
  scale_fill_brewer(type="qual", palette = "Set3") +
  labs(x="Count",
       y="Year",
       color="Cluster") +
  theme_bw()

time_boxplot <- p + 
  geom_boxplot(aes(x=type, y=year, color=type)) +
  geom_jitter(aes(x=type, y=year, color=type)) +
  scale_color_manual(values=c("#66c2a5", "#fc8d62")) +
  labs(x="Measurment",
       y="Time",
       color="Measurement") +
  theme_bw()

```
`cowplot`'s `plot_grid` function takes these plotting objects as input and 
arranges them into a grid.

# Simple compound plots
To generate our first compound figure, we provide the bar and boxplot objects to
`plot_grid`, which arranges them in a single row, next to one another.
```{r}
plot_grid(time_barplot, time_boxplot)
```
## Customising `plot_grid`
`plot_grid` is highly customisable, including layout, labels, alignments and
scaling.

### Labels
To create publication-ready figures, we can set a label for each
plot, by either specifying the desired labels in a vector of the same lengths as
the number of plots (as we do here for 2 plots) or by setting `labels="AUTO"` to
auto-generate upper-case labels or `labels="auto"` to auto-generate lower-case
labels.

```{r}
plot_grid(time_barplot, time_boxplot,
          labels=c(1,2))

```

### Layout
Per default, `plot_grid` chooses a 'one row, two column' layout for the plots
we specified. To specify a 'two row, one column' layout we can either specify the
number of colums `ncol` or number or rows `nrow` argument:

```{r, fig.height=7}
plot_grid(time_barplot, time_boxplot,
          labels=c(1,2),
          nrow=2)
```

### Alignment
You might notice that in the one column layout, the plots are slightly shifted
with respect to their y-axis and legends alignment. Per default, `plot_grid`
keeps the *axis titles* aligned. To align by *actual axis*, we can use a
combination of the `align` and `axis` arguments. Here we want the plots to be
vertically aligned along the left and right margins of the plot panel:

```{r, fig.height=7}
plot_grid(time_barplot, time_boxplot,
          labels=c(1,2),
          nrow=2,
          align="v",
          axis="lr")
```

## Exercises
1. Autogenerate lower-case labels for the compound plot of `time_boxplot` and 
`time_barplot`.
2. Change the default layout by specifying the number of columns. 
3. Read the help function for `plot_grid` and experiment with other `axis`
options.

# Nested compound plots
So far, we have only considered the two time-related plots we generated to
visualise the antigenetic cluster data. In the following few chunks, we will see
how we can add a third plot to the grid that has different dimensions than
then previous two.

## Arranging plots 
Let's start with the default behaviour:
```{r, fig.height=7}
plot_grid(time_barplot, time_boxplot, antigenic_map)
```
This layout does not look great, for several reason:

1. We have a void in the lower right corner;
1. The heights of both rows are the same, which unnessecarily stretches the 
bar and boxplot.

To address both issues, we can arrange a two-column layout, with the first
column containing the antigenic map only, spanning over two rows, and the
bar and boxplot depicted in the first and second row of the second column,
respectively.

To achieve this, we will use two calls to `plot_grid`, the first to generate
our composite for the right-hand column, the second to put this composite 
together with the antigenic map.

```{r, fig.asp = 1}
right_column <- plot_grid(time_boxplot, time_barplot,
                   nrow=2,
                   align="v",
                   axis="lr")

plot_grid(antigenic_map, right_column,
          nrow=1,
          align="h",
          axis="t")
```

## Shared legends
In the visualisation above, we have color-coded both the antigenic map and the
bar plot by cluster. Below, we will have a look how we can add one shared 
legend as an additional row to `plot_grid` and remove the individual legends.
This is what our final figure will look like:

```{r, fig.asp = 1, echo=FALSE}
right_column <- plot_grid( time_boxplot + theme(legend.position = "bottom"),
                           time_barplot + theme(legend.position = "none"),
                           nrow=2,
                           align="v",
                           axis="lr")

top_row <- plot_grid(antigenic_map + theme(legend.position = "none"),
                     right_column, nrow=1)

new_map <- antigenic_map +
  guides(color=guide_legend(nrow=2)) +
  theme(legend.position = "bottom")

cluster_legend <- get_legend(new_map)

plot_grid(top_row, cluster_legend,
          nrow=2,
          rel_heights = c(10,1))

```

Let's have a look at the steps required to get there:

First, we will create the individual components of our panel, this time adding
`legend.position` via `theme`. We create the `right_column`, similar
as above, but, in addition, specify that in this `plot_grid`, the boxplot legend
should move below the plot (`theme(legend.position = "bottom")`) and the legend
from the barplot should be removed with `theme(legend.position = "none")`.
Similarily, we remove the the legend from the `antigenic_map` that we pass to
`plot_grid`.

**Note**: We only move and remove the legends from the objects that we pass to
`plot_grid`, the original objects remain unchanged.


```{r, fig.asp = 1}
right_column <- plot_grid( time_boxplot + theme(legend.position = "bottom"),
                           time_barplot + theme(legend.position = "none"),
                           nrow=2,
                           align="v",
                           axis="lr")

top_row <- plot_grid(antigenic_map + theme(legend.position = "none"),
                     right_column, nrow=1)
```

We then create a new map object from which we extract the
legend. We will place the shared legend as an additional row at the bottom
of the our panel plot, so we want a legend with horizontal layout, as if
it was at the bottom of a plot. Thus, we specify in this new object, that the 
legend should be at the bottom of the plot (as we saw above) and, in addition,
specify that the color legend should have two rows for its elements with 
`guides(color=guide_legend(nrow=2))`.

We then extract the legend from the `new_map` object with `get_legend`:

```{r, fig.asp = 1}
new_map <- antigenic_map +
  guides(color=guide_legend(nrow=2)) +
  theme(legend.position = "bottom")

cluster_legend <- get_legend(new_map)
```

Finally, we will put it all together in a new call to `plot_grid`. In addition
to the arguments that we have already seen for `plot_grid`, we also specify
`rel_heights` here, which says that the ratio of the first row (i.e. our
compound panel) to the second row (the shared legend) should be 10:1.

```{r, fig.asp = 1}
plot_grid(top_row, cluster_legend,
          nrow=2,
          rel_heights = c(10,1))

```

### Exercises
1. What would be a good labeling strategy for this nested plot? Label
accordingly.
2. Play with the ratios provided to `rel_heights`. Why does it not make sense to
provide `rel_widths` in this example? 

# `cowplot` beyond `plot_grid`
Many people come to `cowplot` in the search of a function for composite figures
(`plot_grid`). `cowplot` has other things to offer, most notably for me, it
extends the range of different `themes` provided by `ggplot`. Below an overview
of available cowplot themes, all applied to the `antigenic_map` scatter plot:

```{r, fig.asp = 2.5}
plot_grid(antigenic_map + theme_gray() + ggtitle("Default ggplot"),
          antigenic_map + theme_cowplot()  + ggtitle("theme_cowplot"),
          antigenic_map + theme_half_open()  + ggtitle("theme_half_open"), 
          antigenic_map + theme_minimal_grid() + ggtitle("theme_minimal_grid"),
          antigenic_map + theme_minimal_hgrid() + ggtitle("theme_minimal_hgrid"),
          antigenic_map + theme_minimal_vgrid() + ggtitle("theme_minimal_vgrid"),
          antigenic_map + theme_map() + ggtitle("theme_map"),
          antigenic_map + theme_nothing() + ggtitle("theme_nothing"),
          nrow=4,
          align="vh",
          axis="tblr")
```

All themes are `cowplot` themes apart from the default `ggplot2` theme shown for
reference. The lower right plot shows `theme_nothing` by `cowplot`.

**Note**:
Not all themes are equally well suited for each plot type. The `ggplot2` default
for boxplots chooses a simply horizontal grid for instance, where as the same
theme applied to the scatter plot display horizontal and vertical grids.

## Exercises
1. Discuss which other plot type might benefit from a simpler grid than the
default one chosen?
2. What happens if you add a theme to `plot_grid`?
3. Apply different themes to the three individual plots and visualise in `grid`.


# Additional material
The `ggpubr` package provides some easy-to-use functions for creating and
customizing `ggplot2`- based publication ready plots:
[ggplot2 Based Publication Ready Plots](https://rpkgs.datanovia.com/ggpubr/)
