---
title: "Course evaluation"
author: "Hannah Meyer"
date: "05/02/2020"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache=TRUE,
                      comment = "#>",
                      collapse = TRUE,
                      fig.align = "center",
                      fig.pos = 'h',
                      out.width = "80%")
```


## Read required libraries
```{r}
library(tidyverse)
```

## Read data
Survey data from 28 out of 35 participants of the Workshop:
```{r}
survey <- read_csv("2020_Evaluations_DataVisualR_HMeyer.csv", skip = 2,
                   col_names = FALSE)
```

Format into long format, change type and add short version of questions:
```{r}
tb <- survey[1:5,] %>%
    column_to_rownames("X1") %>%
    t %>%
    as_tibble %>%
    select(-starts_with("Q1")) %>%
    mutate_each(funs(as.integer)) %>%
    pivot_longer(col=starts_with("Q"), names_to = "question_long",
                 values_to="score") %>%
    mutate(question=gsub("Q(\\d).*", "\\1", question_long)) %>%
    mutate(question_long=gsub("Q\\d\\.  (.*)", "\\1", question_long)) %>%
    drop_na
```

## Prepare for plotting per data type
Set separate levels where appropriate, verbalise the selection levels:

```{r}
q3 <- tb %>%
    filter(question == 3) %>% 
    mutate(factorscores=factor(score, levels=1:3,
                               labels=c("too easy", "appropriate", "too hard")))
q2q4 <- tb %>%
    filter(question %in% c(2,4)) %>% 
    mutate(factorscores=factor(score, levels=1:5,
                               labels=c("very poor", "poor", "average",
                                        "good", "very good")))

q5 <- tb %>%
    filter(question == 5) %>% 
    mutate(factorscores=factor(score, levels=1:3,
                               labels=c("very much", "a little", "not at all")))

```

## Visualise results
Show number of survey responses per score and question; summarise similar
questions (Q2 and Q4) into one plot:

```{r}
q2q4_plot <- ggplot(data=q2q4,
                        aes(x=question_long, fill=factorscores))
q2q4_plot <- q2q4_plot + geom_bar(position=position_dodge(preserve = "single")) +
    scale_fill_brewer(type="qual", drop=FALSE) + 
    coord_cartesian(ylim=c(0,23)) +
    labs(x="", 
         y="Score counts",
         fill="Score") +
    theme_bw() +
    theme(legend.position = "bottom")

q3_plot <- ggplot(data=q3,
                        aes(x=question_long, fill=factorscores))
q3_plot <- q3_plot + geom_bar(position=position_dodge(preserve = "single")) +
    scale_fill_manual(values=c( "#d95f02", "#1b9e77","#7570b3"), drop=FALSE) + 
    coord_cartesian(ylim=c(0,23)) +
    labs(x="", 
         y="Score counts",
         fill="Score") +
    theme_bw() +
    theme(legend.position = "bottom")


q5_plot <- ggplot(data=q5,
                        aes(x=question_long, fill=factorscores))
q5_plot <- q5_plot + geom_bar(position=position_dodge(preserve = "single")) +
    scale_fill_manual(values=c("#e7298a","#66a61e", "#e6ab02"), drop=FALSE) + 
    scale_x_discrete(labels="Do you feel more confident in this topic after the workshop?") +
    coord_cartesian(ylim=c(0,23)) +
    labs(x="", 
         y="Score counts",
         fill="Score") +
    theme_bw() +
    theme(legend.position = "bottom")
```

Combine plots in plot grid:
```{r sumary-plot, fig.height=8.5, fig.width=8}
cowplot::plot_grid(q2q4_plot,
                   cowplot::plot_grid(q3_plot, q5_plot, nrow=1), 
                   nrow=2,
                   align="v",
                   axis="r")
```
